Simple repression draft

### 2020-06-04 doc/appendix_01_nonbursty.tex
RP finished appendix

In this section we detail the calculation of mean mRNA levels, fold-changes in
expression, and Fano factors for nonequilibrium promoter models one through four
in Figure~\ref{fig1:means_cartoons}. These are the results that were quoted but
not derived in Sections~\ref{section_02_means} and \ref{sec:beyond_means} of the
main text. In each of these four models, the natural mathematization of their
cartoons is as a chemical master equation such as Eq.~\ref{eq:2state_rep_cme}
for model one. Before jumping into the derivations of the general computation of
the mean mRNA level and the Fano factor we will work through the derivation of an example master equation. In particular we will focus on model 1 from 
Figure~\ref{fig1:means_cartoons}(C). The general steps are applicable to all
other chemical master equations in this work.

@@

\begin{figure}[h!]
\centering
\includegraphics{../figures/si/figS0X_two_state.pdf}
\caption{
\textbf{Two-state promoter chemical master equation.}
(A) Schematic of the two state promoter simple repression model. Rates $k_R^+$
and $k_R^-$ are the association and dissociation rates of the transcriptional
repressor, respectively, $r$ is the transcription initiation rate, and $\gamma$
is the mRNA degradation rate. (B) Schematic depiction of the mRNA count state
transitions. The model in (A) only allows for jumps in mRNA of size 1. The
production of mRNA can only happen when the promoter is in the transcriptionally
active state.}
\label{fig:two_state}
\end{figure}

The chemical master equation describes the continuous time evolution of a
continuous or discrete probability distribution function. In our specific case
we want to describe the time evolution of the discrete mRNA distribution. What
this means is that we want to compute the probability of having $m$ mRNA
molecules at time $t + \Delta t$, where $\Delta t$ is a sufficiently small time
interval such that only one of the possible reactions take place. For the
example that we will workout here in detail we chose the two-state stochastic
simple repression model schematized in Figure~\ref{fig:two_state}(A). To derive
the master equation we will focus more on the representation shown in
Figure~\ref{fig:two_state}(B), where the transitions between different mRNA
counts and promoter states is more explicitly depicted. Given that the DNA
promoter can exist in one of two states -- transcriptionally active state, and
with repressor bound -- we will keep track not only of the mRNA count, but on
which state the promoter is. For this we will keep track of two probability
distributions: The probability of having $m$ mRNAs at time $t$ when the promoter
is in the transcriptionally active state $A$, $p_A(m, t)$, and the equivalent
probability but when the promoter is in the repressor bound state $R$, $p_R(m,
t)$.

Since mRNA production can only take place on the transcriptionally active state
we will focus on this function for our derivation. The repressor bound state
will have an equivalent equation without terms involving the production of
mRNAs. We begin by listing the possible state transition that can happen for a
particular mRNA count with the promoter in the active state. For state changes
in a small time window $\Delta t$ that ``jump into'' state $m$ in the
transcriptionally active state we have
\begin{itemize}
    \item Produce an mRNA, jumping from $m-1$ to $m$.
    \item Degrade an mRNA, jumping from $m+1$ to $m$.
    \item Transition from the repressor bound state $R$ to the active state $A$.
\end{itemize}
Likewise, for state transitions that ``jump out'' of state $m$ in the
transcriptionally inactive state we have
\begin{itemize}
    \item Produce an mRNA, jumping from $m$ to $m+1$.
    \item Degrade an mRNA, jumping from $m$ to $m-1$.
    \item Transition from the active state $A$ to the repressor bound state $R$.
\end{itemize}
The mRNA production does not depend on the current number of mRNAs, therefore
these state transitions happen with probability $r\Delta t$. The same is true
for the promoter state transitions, each happen with probability $k_R^\pm \Delta
t$. As for the mRNA degradation events does must depend on the current number of
mRNA molecules. Each molecule has a constant probability $\gamma \Delta t$ of 
being degraded, so the total probability for an mRNA degradation event to happen
is computed by multiplying this probability by the current number of mRNAs.

To see these terms in action let us compute the probability of having $m$ mRNA
at time $t + \Delta t$ in the transcriptionally active state. This takes the 
form
\begin{equation}
\begin{aligned}
p_A(m, t + \Delta t) &= p_A(m, t) \\
&+ \overbrace{(r\Delta t) p_A(m-1, t)}^{m-1 \rightarrow m}
- \overbrace{(r\Delta t) p_A(m, t)}^{m \rightarrow m+1}\\
% \;\;\;\text{(mRNA production)} \\
&+ \overbrace{(m+1)(\gamma \Delta t) p_A(m+1, t)}^{m+1 \rightarrow m}
- \overbrace{m(\gamma \Delta t) p_A(m, t)}^{m \rightarrow m-1}\\
% \;\;\;\text{(mRNA degradation)} \\
&+\overbrace{(k_R^- \Delta t) p_R(m, t)}^{R \rightarrow A}
-\overbrace{(k_R^+ \Delta t) p_A(m, t)}^{A \rightarrow R},
% \;\;\;\text{(promoter state transition)}
\end{aligned}
\end{equation}
where the overbrace indicates the corresponding state transitions. Notice that
the second to last term on the right-hand side is multiplied by $p_R(m, t)$ 
since the transition from state $R$ to state $A$ depends on the probability of
being in state $R$ to begin with. It is through this term that both probability
distribution functions are coupled. An equivalent equation can be written for
the probability of having $m$ mRNA at time $t + \Delta t$ while in the repressor
bound state, the only difference being that the mRNA production rates are 
removed, and the sign for the promoter state transitions are inverted. This is
\begin{equation}
\begin{aligned}
p_R(m, t + \Delta t) &= p_R(m, t) \\
&+ \overbrace{(m+1)(\gamma \Delta t) p_R(m+1, t)}^{m+1 \rightarrow m}
- \overbrace{m(\gamma \Delta t) p_R(m, t)}^{m \rightarrow m-1}\\
% \;\;\;\text{(mRNA degradation)} \\
&-\overbrace{(k_R^- \Delta t) p_R(m, t)}^{R \rightarrow A}
+\overbrace{(k_R^+ \Delta t) p_A(m, t)}^{A \rightarrow R}.
\end{aligned}
\end{equation}

All we have to do now are simple algebraic steps in order to simplify the 
equations. Let's focus on the transcriptionally active state $A$. First we will
send the term $p_A(m, t)$ to the right-hand side, and then we will divide both
sides of the equation by $\Delta t$. This results in
\begin{equation}
\begin{aligned}
\frac{p_A(m, t + \Delta t) - p_A(m, t)}{\Delta t} &=
r p_A(m-1, t) - r p_A(m, t)\\
&+ (m+1)\gamma p_A(m+1, t)
- m \gamma p_A(m, t)\\
&+k_R^- p_R(m, t)
-k_R^+ p_A(m, t).
\end{aligned}
\end{equation}
Upon taking the limit when $\Delta t \rightarrow 0$ we can transform the 
left-hand side into a derivative, obtaining the chemical master equation
\begin{equation}
\begin{aligned}
\frac{d p_A(m, t)}{dt} &=
r p_A(m-1, t) - r p_A(m, t)\\
&+ (m+1)\gamma p_A(m+1, t)
- m \gamma p_A(m, t)\\
&+k_R^- p_R(m, t)
-k_R^+ p_A(m, t).
\end{aligned}
\end{equation}
Doing equivalent manipulations for the repressor bound state gives an ODE of the
form
\begin{equation}
\begin{aligned}
\frac{d p_R(m, t)}{dt} &=
(m+1)\gamma p_R(m+1, t)
- m \gamma p_R(m, t)\\
&-k_R^- p_R(m, t)
+k_R^+ p_A(m, t).
\end{aligned}
\end{equation}
In the next section we will write these couple ODEs in a more compat form using
matrix notation.

@@

Having derived an example chemical master equation we now focus on writing a
general matrix form for the kinetic models 1-4 shown in
Figure~\ref{fig1:means_cartoons}(C) in the main text. In each of these four
models, the natural mathematization of their cartoons is as a chemical master
equation such as Eq.~\ref{eq:2state_rep_cme} for model one, which for
completeness we reproduce here as
\begin{align}
\begin{split}
\deriv{t}p_R(m,t) =& 
- \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
+ \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_R(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_R(m,t)}^{m \rightarrow m - 1}
\\
\deriv{t}p_U(m,t) =&\; 
\overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
- \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
+ \overbrace{rp_U(m-1,t)}^{m-1 \rightarrow m}
- \overbrace{rp_U(m,t)}^{m \rightarrow m + 1}
\\
&+ \overbrace{(m+1)\gamma p_U(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_U(m,t)}^{m \rightarrow m - 1}.
\label{eq:poisson_promoter_cme_appdx}
\end{split}
\end{align}
Here $p_R(m,t)$ and $p_U(m,t)$ are the probabilities of finding the system with
$m$ mRNA molecules at time $t$ either in the repressor bound or unbound states,
respectively. $r$ is the probability per unit time that a transcript will be
initiated when the repressor is unbound, and $\gamma$ is the probability per
unit time for a given mRNA to be degraded. $k_R^-$ is the probability per unit
time that a bound repressor will unbind, while $k_R^+$ is the probability per
unit time that an unbound operator will become bound by a repressor. Assuming
mass action kinetics, $k_R^+$ is proportional to repressor copy number $R$.

Next consider the cartoon for nonequilibrium model two in
Figure~\ref{fig1:means_cartoons}(C). Now we must track probabilities $p_R$,
$p_P$, and $p_E$ for the repressor bound, empty, and polymerase bound states,
respectively. By analogy to~\ref{eq:poisson_promoter_cme_appdx}, the master
equation for model two can be written
\begin{align}
\begin{split}
\deriv{t}p_R(m,t) =& 
- \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
+ \overbrace{k_R^+ p_E(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_R(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_R(m,t)}^{m \rightarrow m - 1}
\\
\deriv{t}p_E(m,t) =&\; 
  \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
- \overbrace{k_R^+ p_E(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_E(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_E(m,t)}^{m \rightarrow m - 1}.
\\
&
+ \overbrace{k_P^- p_P(m,t)}^{A \rightarrow U}
- \overbrace{k_P^+ p_E(m,t)}^{U \rightarrow A}
+ \overbrace{rp_P(m-1,t)}^{m-1 \rightarrow m,\;A \rightarrow U}
\\
\deriv{t}p_P(m,t) =&\; 
- \overbrace{k_P^- p_P(m,t)}^{A \rightarrow U}
+ \overbrace{k_P^+ p_E(m,t)}^{U \rightarrow A}
+ \overbrace{(m+1)\gamma p_P(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_P(m,t)}^{m \rightarrow m - 1}.
\\
&- \overbrace{rp_P(m,t)}^{m \rightarrow m + 1,\;A \rightarrow U}.
\label{eq:model2_cme_appdx}
\end{split}
\end{align}
$k_P^+$ and $k_P^-$ are defined in close analogy to $k_R^+$ and $k_R^-$, except
for RNAP binding and unbinding instead of repressor. Similarly $p_P(m,t)$ is
defined for the active (RNAP-bound) state exactly as are $p_R(m,t)$ and
$p_E(m,t)$ for the repressor bound and unbound states, respectively. The new
subtlety Eq.~\ref{eq:model2_cme_appdx} introduces compared to
Eq.~\ref{eq:poisson_promoter_cme_appdx} is that when mRNAs are produced, the
promoter state also changes. This is encoded by the terms involving $r$, the
last term in each of the equations for $p_E$ and $p_P$. The former accounts for
arrivals in the unbound state and the latter accounts for departures from the
RNAP-bound state.

@@

\subsection{General forms for mean mRNA and Fano factor}
Our task now is to derive expressions for the first two moments of the
steady-state mRNA distribution from Eq.~\ref{eq:generic_cme_appdx}. Our
treatment of this is analogous to that given in~\cite{Sanchez2011}
and~\cite{Razo-Mejia2020}. We first obtain the steady-state limit of
Eq.~\ref{eq:generic_cme_appdx} in which the time derivative vanishes, giving
\begin{equation}
0 =
\left( \mathbf{K} - \mathbf{R_D} - \gamma m \mathbf{I} \right) \vec{p}(m)
                + \mathbf{R_A} \vec{p}(m-1) +
                \gamma (m+1) \mathbf{I} \vec{p}(m+1),
\label{eq:generic_cme_ss}
\end{equation}
From this, we want to compute
\begin{equation}
\langle{m}\rangle = \sum_S \sum_{m=0}^\infty m \, p_S(m)
\end{equation}
and
\begin{equation}
\langle{m^2}\rangle = \sum_S \sum_{m=0}^\infty m^2 p_S(m)
\end{equation}
which define the average values of $m$ and $m^2$ at steady state, where the
averaging is over all possible mRNA copy numbers and promoter states $S$. For
example, for model one in Figure~\ref{fig1:means_cartoons}(C), the sum on $S$
would cover repressor bound and unbound states ($R$ and $U$ respectively), for
model two, the sum would cover repressor bound, polymerase bound, and empty
states ($R$, $P$, and $E$), and so on for the other models.

@@

Next we can find the mean mRNA copy number $\langle{m}\rangle$ from
$\langle\vec{m}\rangle$ according to
\begin{equation}
\langle{m}\rangle = \vec{1}^\dagger\cdot\langle\vec{m}\rangle,
\label{eq:m_from_vecm}
\end{equation}
where $\vec{1}^\dagger$ is a row vector whose elements are all 1.
Eq.~\ref{eq:m_from_vecm} holds since the $i$-th element of the column vector
$\langle\vec{m}\rangle$ is the mean mRNA value conditional on the system
occupying the $i-$th promoter state, so the dot product with $\vec{1}^\dagger$
amounts to simply summing the elements of $\langle\vec{m}\rangle$. Rather than
solving Eq.~\ref{eq:generic_cme_deriv_040} for $\langle\vec{m}\rangle$ and then
computing $\vec{1}^\dagger\cdot\langle\vec{m}\rangle$, we may take a shortcut by
multiplying Eq.~\ref{eq:generic_cme_deriv_040} by $\vec{1}^\dagger$ first. The
key observation that makes this useful is that
\begin{equation}
\vec{1}^\dagger \cdot (\mathbf{K} - \mathbf{R_D} + \mathbf{R_A}) = 0.
\end{equation}
Intuitively, this equality holds because each column of this matrix
represents transitions to and from a given promoter state. In any given column,
the diagonal encodes all departing transitions and off-diagonals encode all
arriving transitions. Conservation of probability means that each column
sums to zero, and summing columns is exactly the operation that multiplying by
$\vec{1}^\dagger$ carries out.

>>>>

In this section we detail the calculation of mean mRNA levels, fold-changes in
expression, and Fano factors for nonequilibrium promoter models one through four
in Figure~\ref{fig1:means_cartoons}. These are the results that were quoted but
not derived in Sections~\ref{section_02_means} and \ref{sec:beyond_means} of the
main text. In each of these four models, the natural mathematicization of their
cartoons is as a chemical master equation such as Eq.~\ref{eq:2state_rep_cme}
for model one. Before jumping into the derivations of the general computation of
the mean mRNA level and the Fano factor we will work through the derivation of an example master equation. In particular we will focus on model 1 from 
Figure~\ref{fig1:means_cartoons}(C). The general steps are applicable to all
other chemical master equations in this work.

@@

\begin{figure}[h!]
\centering
\includegraphics{../figures/si/figS0X_two_state.pdf}
\caption{
\textbf{Two-state promoter chemical master equation.}
(A) Schematic of the two state promoter simple repression model. Rates $k_R^+$
and $k_R^-$ are the association and dissociation rates of the transcriptional
repressor, respectively, $r$ is the transcription initiation rate, and $\gamma$
is the mRNA degradation rate. (B) Schematic depiction of the mRNA count state
transitions. The model in (A) only allows for jumps in mRNA of size 1. The
production of mRNA can only occur when the promoter is in the transcriptionally
active state.}
\label{fig:two_state}
\end{figure}

The chemical master equation describes the continuous time evolution of a
continuous or discrete probability distribution function. In our specific case
we want to describe the time evolution of the discrete mRNA distribution. What
this means is that we want to compute the probability of having $m$ mRNA
molecules at time $t + \Delta t$, where $\Delta t$ is a sufficiently small time
interval such that only one of the possible reactions take place during that time interval. For the
example that we will work out here in detail we chose the two-state stochastic
simple repression model schematized in Figure~\ref{fig:two_state}(A). To derive
the master equation we will focus more on the representation shown in
Figure~\ref{fig:two_state}(B), where the transitions between different mRNA
counts and promoter states is more explicitly depicted. Given that the DNA
promoter can exist in one of two states -- transcriptionally active state, and
with repressor bound -- we will keep track not only of the mRNA count, but on
which state the promoter is. For this we will keep track of two probability
distributions: The probability of having $m$ mRNAs at time $t$ when the promoter
is in the transcriptionally active state $A$, $p_A(m, t)$, and the equivalent
probability but when the promoter is in the repressor bound state $R$, $p_R(m,
t)$.

Since mRNA production can only take place in the transcriptionally active state
we will focus on this function for our derivation. The repressor bound state
will have an equivalent equation without terms involving the production of
mRNAs. We begin by listing the possible state transition that can happen for a
particular mRNA count with the promoter in the active state. For state changes
in a small time window $\Delta t$ that ``jump into'' state $m$ in the
transcriptionally active state we have
\begin{itemize}
    \item Produce an mRNA, jumping from $m-1$ to $m$.
    \item Degrade an mRNA, jumping from $m+1$ to $m$.
    \item Transition from the repressor bound state $R$ with m mRNAs  to the active state $A$
    with m mRNAs.
\end{itemize}
Likewise, for state transitions that ``jump out'' of state $m$ in the
transcriptionally inactive state we have
\begin{itemize}
    \item Produce an mRNA, jumping from $m$ to $m+1$.
    \item Degrade an mRNA, jumping from $m$ to $m-1$.
    \item Transition from the active state $A$ with m mRNAs to the repressor bound state $R$ with
    m mRNAs.
\end{itemize}
The mRNA production does not depend on the current number of mRNAs, therefore
these state transitions occur with probability $r\Delta t$. The same is true
for the promoter state transitions, each occurs with probability $k_R^\pm \Delta
t$. As for the mRNA degradation events, these transitions depend on the current number of
mRNA molecules since the more molecules of mRNA there are, the more will decay during
a given time interval. Each molecule has a constant probability $\gamma \Delta t$ of 
being degraded, so the total probability for an mRNA degradation event to happen
is computed by multiplying this probability by the current number of mRNAs.

To see these terms in action let us compute the probability of having $m$ mRNA
at time $t + \Delta t$ in the transcriptionally active state. This takes the 
form
\begin{equation}
\begin{aligned}
p_A(m, t + \Delta t) &= p_A(m, t) \\
&+ \overbrace{(r\Delta t) p_A(m-1, t)}^{m-1 \rightarrow m}
- \overbrace{(r\Delta t) p_A(m, t)}^{m \rightarrow m+1}\\
% \;\;\;\text{(mRNA production)} \\
&+ \overbrace{(m+1)(\gamma \Delta t) p_A(m+1, t)}^{m+1 \rightarrow m}
- \overbrace{m(\gamma \Delta t) p_A(m, t)}^{m \rightarrow m-1}\\
% \;\;\;\text{(mRNA degradation)} \\
&+\overbrace{(k_R^- \Delta t) p_R(m, t)}^{R \rightarrow A}
-\overbrace{(k_R^+ \Delta t) p_A(m, t)}^{A \rightarrow R},
% \;\;\;\text{(promoter state transition)}
\end{aligned}
\end{equation}
where the overbrace indicates the corresponding state transitions. Notice that
the second to last term on the right-hand side is multiplied by $p_R(m, t)$ 
since the transition from state $R$ to state $A$ depends on the probability of
being in state $R$ to begin with. It is through this term that the dynamics of the two probability
distribution functions ($p_R(m,t)$ and $p_A(m,t)$ are coupled. An equivalent equation can be written for
the probability of having $m$ mRNA at time $t + \Delta t$ while in the repressor
bound state, the only difference being that the mRNA production rates are 
removed, and the sign for the promoter state transitions are inverted. This is
\begin{equation}
\begin{aligned}
p_R(m, t + \Delta t) &= p_R(m, t) \\
&+ \overbrace{(m+1)(\gamma \Delta t) p_R(m+1, t)}^{m+1 \rightarrow m}
- \overbrace{m(\gamma \Delta t) p_R(m, t)}^{m \rightarrow m-1}\\
% \;\;\;\text{(mRNA degradation)} \\
&-\overbrace{(k_R^- \Delta t) p_R(m, t)}^{R \rightarrow A}
+\overbrace{(k_R^+ \Delta t) p_A(m, t)}^{A \rightarrow R}.
\end{aligned}
\end{equation}

All we have to do now are simple algebraic steps in order to simplify the 
equations. Let's focus on the transcriptionally active state $A$. First we will
send the term $p_A(m, t)$ to the right-hand side, and then we will divide both
sides of the equation by $\Delta t$. This results in
\begin{equation}
\begin{aligned}
\frac{p_A(m, t + \Delta t) - p_A(m, t)}{\Delta t} &=
r p_A(m-1, t) - r p_A(m, t)\\
&+ (m+1)\gamma p_A(m+1, t)
- m \gamma p_A(m, t)\\
&+k_R^- p_R(m, t)
-k_R^+ p_A(m, t).
\end{aligned}
\end{equation}
Upon taking the limit when $\Delta t \rightarrow 0$ we can transform the 
left-hand side into a derivative, obtaining the chemical master equation
\begin{equation}
\begin{aligned}
\frac{d p_A(m, t)}{dt} &=
r p_A(m-1, t) - r p_A(m, t)\\
&+ (m+1)\gamma p_A(m+1, t)
- m \gamma p_A(m, t)\\
&+k_R^- p_R(m, t)
-k_R^+ p_A(m, t).
\end{aligned}
\end{equation}
Doing equivalent manipulations for the repressor bound state gives an ODE of the
form
\begin{equation}
\begin{aligned}
\frac{d p_R(m, t)}{dt} &=
(m+1)\gamma p_R(m+1, t)
- m \gamma p_R(m, t)\\
&-k_R^- p_R(m, t)
+k_R^+ p_A(m, t).
\end{aligned}
\end{equation}
In the next section we will write these coupled ODEs in a more compact form using
matrix notation.

@@

Having derived an example chemical master equation we now focus on writing a
general matrix form for the kinetic models 1-4 shown in
Figure~\ref{fig1:means_cartoons}(C) in the main text. In each of these four
models, the natural mathematicization of their cartoons is as a chemical master
equation such as Eq.~\ref{eq:2state_rep_cme} for model one, which for
completeness we reproduce here as
\begin{align}
\begin{split}
\deriv{t}p_R(m,t) =& 
- \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
+ \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_R(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_R(m,t)}^{m \rightarrow m - 1}
\\
\deriv{t}p_U(m,t) =&\; 
\overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
- \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
+ \overbrace{rp_U(m-1,t)}^{m-1 \rightarrow m}
- \overbrace{rp_U(m,t)}^{m \rightarrow m + 1}
\\
&+ \overbrace{(m+1)\gamma p_U(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_U(m,t)}^{m \rightarrow m - 1}.
\label{eq:poisson_promoter_cme_appdx}
\end{split}
\end{align}
Here $p_R(m,t)$ and $p_U(m,t)$ are the probabilities of finding the system with
$m$ mRNA molecules at time $t$ either in the repressor bound or unbound states,
respectively. $r$ is the probability per unit time that a transcript will be
initiated when the repressor is unbound, and $\gamma$ is the probability per
unit time for a given mRNA to be degraded. $k_R^-$ is the probability per unit
time that a bound repressor will unbind, while $k_R^+$ is the probability per
unit time that an unbound operator will become bound by a repressor. Assuming
mass action kinetics, $k_R^+$ is proportional to repressor copy number $R$.

Next consider the cartoon for nonequilibrium model two in
Figure~\ref{fig1:means_cartoons}(C). Now we must track probabilities $p_R$,
$p_P$, and $p_E$ for the repressor bound, empty, and polymerase bound states,
respectively. By analogy to eqn.~\ref{eq:poisson_promoter_cme_appdx}, the master
equation for model two can be written
\begin{align}
\begin{split}
\deriv{t}p_R(m,t) =& 
- \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
+ \overbrace{k_R^+ p_E(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_R(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_R(m,t)}^{m \rightarrow m - 1}
\\
\deriv{t}p_E(m,t) =&\; 
  \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
- \overbrace{k_R^+ p_E(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_E(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_E(m,t)}^{m \rightarrow m - 1}.
\\
&
+ \overbrace{k_P^- p_P(m,t)}^{A \rightarrow U}
- \overbrace{k_P^+ p_E(m,t)}^{U \rightarrow A}
+ \overbrace{rp_P(m-1,t)}^{m-1 \rightarrow m,\;A \rightarrow U}
\\
\deriv{t}p_P(m,t) =&\; 
- \overbrace{k_P^- p_P(m,t)}^{A \rightarrow U}
+ \overbrace{k_P^+ p_E(m,t)}^{U \rightarrow A}
+ \overbrace{(m+1)\gamma p_P(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_P(m,t)}^{m \rightarrow m - 1}.
\\
&- \overbrace{rp_P(m,t)}^{m \rightarrow m + 1,\;A \rightarrow U}.
\label{eq:model2_cme_appdx}
\end{split}
\end{align}
$k_P^+$ and $k_P^-$ are defined in close analogy to $k_R^+$ and $k_R^-$, except
for RNAP binding and unbinding instead of repressor. Similarly $p_P(m,t)$ is
defined for the active (RNAP-bound) state exactly as are $p_R(m,t)$ and
$p_E(m,t)$ for the repressor bound and unbound states, respectively. The new
subtlety Eq.~\ref{eq:model2_cme_appdx} introduces compared to
Eq.~\ref{eq:poisson_promoter_cme_appdx} is that when mRNAs are produced, the
promoter state also changes. This is encoded by the terms involving $r$, the
last term in each of the equations for $p_E$ and $p_P$. The former accounts for
arrivals in the unbound state and the latter accounts for departures from the
RNAP-bound state.

@@

\subsection{General forms for mean mRNA and Fano factor}
Our task now is to derive expressions for the first two moments of the
steady-state mRNA distribution from Eq.~\ref{eq:generic_cme_appdx}. Our
treatment of this is analogous to that given in Refs.~\cite{Sanchez2011}
and~\cite{Razo-Mejia2020}. We first obtain the steady-state limit of
Eq.~\ref{eq:generic_cme_appdx} in which the time derivative vanishes, giving
\begin{equation}
0 =
\left( \mathbf{K} - \mathbf{R_D} - \gamma m \mathbf{I} \right) \vec{p}(m)
                + \mathbf{R_A} \vec{p}(m-1) +
                \gamma (m+1) \mathbf{I} \vec{p}(m+1),
\label{eq:generic_cme_ss}
\end{equation}
From this, we want to compute
\begin{equation}
\langle{m}\rangle = \sum_S \sum_{m=0}^\infty m \, p_S(m)
\end{equation}
and
\begin{equation}
\langle{m^2}\rangle = \sum_S \sum_{m=0}^\infty m^2 p_S(m)
\end{equation}
which define the average values of $m$ and $m^2$ at steady state, where the
averaging is over all possible mRNA copy numbers and promoter states $S$. For
example, for model one in Figure~\ref{fig1:means_cartoons}(C), the sum on $S$
would cover repressor bound and unbound states ($R$ and $U$ respectively), for
model two, the sum would cover repressor bound, polymerase bound, and empty
states ($R$, $P$, and $E$), and so on for the other models.

@@

Next we can find the mean mRNA copy number $\langle{m}\rangle$ from
$\langle\vec{m}\rangle$ according to
\begin{equation}
\langle{m}\rangle = \vec{1}^\dagger\cdot\langle\vec{m}\rangle,
\label{eq:m_from_vecm}
\end{equation}
where $\vec{1}^\dagger$ is a row vector whose elements are all 1.
Eq.~\ref{eq:m_from_vecm} holds since the $i^{th}$ element of the column vector
$\langle\vec{m}\rangle$ is the mean mRNA value conditional on the system
occupying the $i^{th}$ promoter state, so the dot product with $\vec{1}^\dagger$
amounts to simply summing the elements of $\langle\vec{m}\rangle$. Rather than
solving Eq.~\ref{eq:generic_cme_deriv_040} for $\langle\vec{m}\rangle$ and then
computing $\vec{1}^\dagger\cdot\langle\vec{m}\rangle$, we may take a shortcut by
multiplying Eq.~\ref{eq:generic_cme_deriv_040} by $\vec{1}^\dagger$ first. The
key observation that makes this useful is that
\begin{equation}
\vec{1}^\dagger \cdot (\mathbf{K} - \mathbf{R_D} + \mathbf{R_A}) = 0.
\end{equation}
Intuitively, this equality holds because each column of this matrix
represents transitions to and from a given promoter state. In any given column,
the diagonal encodes all departing transitions and off-diagonals encode all
arriving transitions. Conservation of probability means that each column
sums to zero, and summing columns is exactly the operation that multiplying by
$\vec{1}^\dagger$ carries out.
